---
title: "Final Project"
author: "Jai abhishek singh"
date: "2024-05-01"
output:
  html_document: default
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
---


## Exploring Sonic Success: A Deep Dive into Spotify Songs and Their Popularity

### What is Spotify?

Launched in 2008, Spotify has evolved into an integral component of the contemporary music ecosystem, boasting an extensive catalog comprising over 70 million tracks and serving the varied preferences of its 345 million active users globally.  
  
 Spotify employs a distinctive streaming model that grants users access to an expansive repertoire of songs without the encumbrance of traditional downloads. Its algorithmic sophistication is instrumental in curating playlists and recommendations tailored to individual user preferences, thereby fashioning a bespoke auditory experience. This democratization of music has fundamentally altered the dynamics of how we engage with diverse songs and genres, fostering a global community of impassioned music enthusiasts.  
  
  
### Focus of this Study 

The research inquiry of this study engenders a comprehensive analysis of Spotify's expansive dataset. The objective is to decipher discernible patterns and constituent elements that propel specific musical compositions into the spotlight. In navigating this sonic realm, our investigative endeavor adheres rigorously to academic standards, driven by an earnest quest for insights into the transformative influence of the contemporary digital age on the dynamics of music popularity.  
  
The central theme guiding our inquiry revolves around the exploration of "What attributes contribute to the popularity of a song or genre?" This investigation is underpinned by a meticulously curated dataset obtained from Kaggle. Leveraging this dataset, our aim is to delve into specific subquestions that will collectively culminate in a comprehensive answer to our primary inquiry.  
  

## Exploring the Dataset

we dive into what the dataset encompasses and how the valuable information it contains was utilized in our study.

### What Does it Contain?

The dataset that has been selected from Kaggle is a rich collection of information about 30,000 songs available on Spotify. Each song serves as an individual case within the dataset, contributing to a comprehensive overview of musical characteristics. The dataset encompasses a total of 24 variables, providing a multifaceted description of each song. Notably, 10 of these variables are characterized by characters, 5 are represented as integers, and the remaining 9 are of the number type.  
  
Among the 24 variables, a careful selection process has been undertaken to focus on the most relevant and useful features. The chosen variables include essential aspects such as popularity, genre, sub-genre, and release dates. These variables are integral for gaining insights into the songs' overall appeal, categorization, and temporal patterns. By narrowing down the dataset to these key variables, you can streamline your analysis and concentrate on the most meaningful aspects of the music data.  
  
To better understand the nature of the selected variables, it's crucial to delve into their measurement scales. For instance, popularity is gauged on a scale of 0 to 100, reflecting the enduring appeal of each song over time. Meanwhile, other features like danceability, energy, speechiness, liveness, and acousticness are measured on a scale from 0 to 1, providing nuanced insights into the musical composition. The loudness variable, measured in decibels, spans from -60 to 0, offering a quantitative assessment of the songs' audio intensity. Additionally, the binary variable, mode, serves as a categorical indicator, distinguishing between major keys (represented as 1) and minor keys (represented as 0).  
  
In summary, the dataset holds a diverse array of information about songs on Spotify, with a thoughtful emphasis on key variables that offer a comprehensive understanding of each musical piece. The varied measurement scales of these variables add depth to your analysis, enabling you to explore the intricacies of musical characteristics and trends within the dataset.  
  
### Data Pre-processing
  
Managing a dataset of 30,000 Spotify songs came with its challenges. Firstly, some songs appeared multiple times due to being in different playlists and genres. To be fair, i kept all instances of repeated songs, ensuring they were considered separately for each genre they associated with.  
  
Another challenge was the inconsistent date formats for song releases. To make things clearer, i created a new column called "Cleaned Up Release Date," focusing only on the year of release. This helped maintain consistency and simplicity in analyzing when the songs were released.  
  
When it came to visualizing the data, dealing with individual release years proved messy. So, i decided to group the years into decades, making it easier to see trends. To keep things balanced, we excluded songs from 1957 to 1959 and those from 2020, as the data for these years was comparatively limited. This approach allowed to present a more straightforward and meaningful picture of how music has evolved over broader timeframes.  
  
This made it much easier when it came to creating a summary table to understand the trends of each decade when it came to popularity of songs released in that decade.  
  

```{r frequencyTableDecade,echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Install and load required packages
packages_to_install <- c("tidyverse", "dplyr", "ggplot2", "lubridate", "knitr", "kableExtra")

for (package in packages_to_install) {
  if (!requireNamespace(package, quietly = TRUE)) {
    install.packages(package)
  }
}

# Load libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)
library(knitr)
library(kableExtra)  # Load kableExtra package

# Load data from GitHub
github_url <- "https://raw.githubusercontent.com/jailancer/final-project-jailancer/main/Spotify%20Songs%20NEW.csv"
spotify_data <- read.csv(github_url)

# Statistical Chart for Decades
filtered_data_decades <- spotify_data %>%
  filter(Cleaned.Up.Release.Date > 1960 & Cleaned.Up.Release.Date < 2020)

summary_stats_decades <- filtered_data_decades %>%
  mutate(Decade = cut(Cleaned.Up.Release.Date, breaks = seq(1960, 2020, by = 10),
                      labels = c("1960s", "1970s", "1980s", "1990s", "2000s", "2010s"))) %>%
  group_by(Decade) %>%
  summarise(
    n = n(),
    min = min(track_popularity),
    Q1 = quantile(track_popularity, probs = 0.25),
    Median = median(track_popularity),
    Q3 = quantile(track_popularity, probs = 0.75),
    Max = max(track_popularity)
  ) %>%
  ungroup()

# Print nicely formatted table with center alignment and caption
kable(summary_stats_decades, "latex", booktabs = TRUE, escape = FALSE, align = "c", caption = "Decades Frequency Table")

```
  
The Decade Frequency table shown here shows the statistics of the popularity of songs in each decade. This information is important for tracking shifts in musical preferences over the decades and understanding what appealed to audiences in different eras.


  
## Research Questions
  
The overarching theme is to find out what makes a song on Spotify popular. To analyse this, i formed a few subquestions to get down to the root cause.

### What are the most and least popular genres in each decade?
  
With this question, i can creat a graph that easily shows the most and least popular genres in each decade. One major observation  which is noticed was that the 1960s had the most and one of the least average popularity of songs overall. The most popular genre was Pop with Latin second. Latin and Pop have similar audiences so it is safe to say that the most popular songs are going to be in one of these two categories. If a song wanted to be a popular song, it would be a good idea to make it for these type of audience. Even with Latin and Pop being extremely popular the most popular genres can switch like in the 1980s and 1990s. Their most popular genres was rock. This means that even though Latin and Pop are very popular, the popularity can change with the decades. That would mean that if a song wanted to be popular a good idea would be to follow the trend of the decade.  
  
```{r genreBarChart, echo=FALSE, warning=FALSE, message=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com"))

# Install and load required packages
packages_to_install <- c("tidyverse", "dplyr", "ggplot2", "lubridate")

for (package in packages_to_install) {
  if (!requireNamespace(package, quietly = TRUE)) {
    install.packages(package)
  }
}

# Load libraries
library(tidyverse)
library(dplyr)
library(ggplot2)
library(lubridate)


# Create a decade variable with custom labels
spotify_data$Decade <- cut(spotify_data$Cleaned.Up.Release.Date, breaks = seq(1950, 2020, by = 10),
                           labels = c("1950s", "1960s", "1970s", "1980s", "1990s", "2000s", "2010s"))

# Create a summary table with average popularity for each genre in each decade
popularity_summary <- spotify_data %>%
  group_by(Decade, playlist_genre) %>%
  summarise(Avg_Popularity = mean(track_popularity))

# Identify the highest and lowest popularity genre for each decade
top_genre <- popularity_summary %>%
  group_by(Decade) %>%
  filter(Avg_Popularity == max(Avg_Popularity))

bottom_genre <- popularity_summary %>%
  group_by(Decade) %>%
  filter(Avg_Popularity == min(Avg_Popularity))

# Combine the highest and lowest popularity genres
top_bottom_genres <- bind_rows(top_genre, bottom_genre)

# Exclude "1950s" and "N/A" from the data
top_bottom_genres <- top_bottom_genres %>%
  filter(Decade %in% c("1960s", "1970s", "1980s", "1990s", "2000s", "2010s"))

# Plot the bar chart
ggplot(top_bottom_genres, aes(x = Decade, y = Avg_Popularity, fill = playlist_genre)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8), width = 0.7) +
  labs(title = "Highest and Lowest Popularity Genre for Each Decade",
       x = "Decade",
       y = "Average Popularity") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set3")

```
  
### Question 2 : How have factors like danceability, loudness, duration, and tempo affected popularity over the years?
  

Tolook at these factors we can make graphs that show the 1960s vs the 2010s with danceability, loudness, duration, and tempo vs popularity. With these graphs we can see the trend of the decade and compare it with the other decade's trend. This will provide us with the knowledge to make assumptions with what factors in a song will increase it's popularity.  

  
##### Little to no Change

The graphs of loudness and tempo had similar trend lines.  

##### Opposite Change

The only factor that had a opposite change was the danceability vs popularity graph.  

##### Same Change

The graphs that had a big change but not in direction was the duration vs popularity. 

  
```{r popularityScatterPlots, echo=FALSE, message=FALSE, warning=FALSE}
options(repos = c(CRAN = "https://cran.rstudio.com"))

# Create a decade variable with custom labels
spotify_data$Decade <- cut(spotify_data$Cleaned.Up.Release.Date, breaks = seq(1950, 2020, by = 10),
                           labels = c("1950s", "1960s", "1970s", "1980s", "1990s", "2000s", "2010s"))

# Filter data for the 1960s decade
songs_1960s <- spotify_data %>%
  filter(Decade == "1960s")

# Filter data for the 2010s decade
songs_2010s <- spotify_data %>%
  filter(Decade == "2010s")

# Danceability 1960
ggplot(songs_1960s, aes(x = danceability, y = track_popularity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  
  labs(title = "Scatterplot of Danceability vs Popularity (1960s)",
       x = "Danceability",
       y = "Popularity") +
  theme_minimal()


# Danceability 2010
ggplot(songs_2010s, aes(x = danceability, y = track_popularity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  labs(title = "Scatterplot of Danceability vs Popularity (2010s)",
       x = "Danceability",
       y = "Popularity") +
  theme_minimal()


# Loudness 1960
ggplot(songs_1960s, aes(x = loudness, y = track_popularity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  
  labs(title = "Scatterplot of Loudness vs Popularity (1960s)",
       x = "Loudness",
       y = "Popularity") +
  theme_minimal()


# Loudness 2010
ggplot(songs_2010s, aes(x = loudness, y = track_popularity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  labs(title = "Scatterplot of Loudness vs Popularity (2010s)",
       x = "Loudness",
       y = "Popularity") +
  theme_minimal()


# Duration 1960
ggplot(songs_1960s, aes(x = duration_ms, y = track_popularity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") + 
  labs(title = "Scatterplot of Duration vs Popularity (1960s)",
       x = "Duration (ms)",
       y = "Popularity") +
  theme_minimal()


# Duration 2010
ggplot(songs_2010s, aes(x = duration_ms, y = track_popularity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  labs(title = "Scatterplot of Duration vs Popularity (2010s)",
       x = "Duration (ms)",
       y = "Popularity") +
  theme_minimal()


# Tempo 1960
ggplot(songs_1960s, aes(x = tempo, y = track_popularity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +  
  labs(title = "Scatterplot of Tempo vs Popularity (1960s)",
       x = "Tempo",
       y = "Popularity") +
  theme_minimal()


# Tempo 2010
ggplot(songs_2010s, aes(x = tempo, y = track_popularity)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red") +  
  labs(title = "Scatterplot of Tempo vs Popularity (2010s)",
       x = "Tempo",
       y = "Popularity") +
  theme_minimal()
```
  
## Conclusion
  
These Spotify songs have many different factors that contribute to their popularity. Although this database covers a lot of factors for these songs, it does not cover nearly all of the attributes like audience or song artist. With this database, one major contributor is the audience's preference. as we can see with the most/least popular genre graph that the most popular genres would change over the decades. This is because of many factors but a leading one would be the audience's preferences. The audience can change what they like over the decades and what they consider to be hot in the decade.
If a song wants to be popular it must keep up with what is a new trend and most follow the trend of the decade.
  
